---
title: "Sri Lanka Dengue Surveillance Dashboard"
format:
  dashboard:
    css: styles.css
editor: source
author: "Madhushani Perera"
date: today
theme: cosmo
execute:
  cache: true
  freeze: auto
  warning: false
  message: false
---

```{=html}
<style>
/* Push navbar author onto its own line, well below the title  */
#quarto-dashboard-header .navbar-brand-container {
  flex-wrap: wrap;
  align-items: flex-start;
}
#quarto-dashboard-header .navbar-title {
  flex-basis: 100%;
}
#quarto-dashboard-header .navbar-author {
  flex-basis: 100%;
  display: block;
  margin-top: 10px;
  padding-top: 4px;
  font-size: 0.72em;
  opacity: 0.75;
  order: 3;
}

/*  Center value box content (icon + title + value), both ways  */
/* Force flexbox centering at every level of the value box so text can't
   end up bottom-anchored regardless of bslib's default grid/height rules */
.bslib-value-box.card,
.value-box.card {
  display: flex !important;
}
.bslib-value-box .card-body,
.value-box .card-body {
  display: flex !important;
  align-items: center !important;
  height: 100% !important;
  width: 100% !important;
}
.bslib-value-box .value-box-grid,
.value-box .value-box-grid {
  display: flex !important;
  flex-direction: row !important;
  align-items: center !important;
  justify-content: center !important;
  height: 100% !important;
  width: 100% !important;
  gap: 1rem;
}
.bslib-value-box .value-box-showcase,
.value-box .value-box-showcase {
  display: flex !important;
  align-items: center !important;
  justify-content: center !important;
  height: auto !important;
}
.bslib-value-box .value-box-area,
.value-box .value-box-area {
  display: flex !important;
  flex-direction: column !important;
  align-items: center !important;
  justify-content: center !important;
  text-align: center !important;
  height: auto !important;
}
.bslib-value-box .value-box-title,
.bslib-value-box .value-box-value,
.value-box .value-box-title,
.value-box .value-box-value {
  text-align: center !important;
  width: 100%;
  margin: 0 !important;
}
</style>
```

```{=html}
<link rel="stylesheet" href="https://cdn.jsdelivr.net/npm/flatpickr/dist/flatpickr.min.css">
<script src="https://cdn.jsdelivr.net/npm/flatpickr"></script>
```

```{r}
#| label: load-packages
#| cache: false

# Set JAVA_HOME BEFORE rJava loads (required on Windows)
if (Sys.getenv("JAVA_HOME") == "") {
  # 1. Try locating java.exe via PATH
  java_bin <- Sys.which("java")
  if (nchar(java_bin) > 0) {
    Sys.setenv(JAVA_HOME = dirname(dirname(normalizePath(java_bin,
                                                         winslash = "/"))))
  } else if (.Platform$OS.type == "windows") {
    # 2. Windows fallback: search common JDK/JRE install locations
    candidates <- c(
      list.files("C:/Program Files/Java",         full.names = TRUE),
      list.files("C:/Program Files/Eclipse Adoptium", full.names = TRUE),
      list.files("C:/Program Files/Microsoft",    full.names = TRUE,
                 pattern = "jdk|jre"),
      list.files("C:/Program Files/OpenJDK",      full.names = TRUE)
    )
    # Keep only dirs that contain bin/java.exe
    valid <- candidates[file.exists(file.path(candidates, "bin/java.exe"))]
    if (length(valid) > 0) {
      Sys.setenv(JAVA_HOME = normalizePath(valid[1], winslash = "/"))
    } else {
      stop(paste0(
        "Java not found. Please install a JDK (e.g. from https://adoptium.net) ",
        "and either set JAVA_HOME in your environment or add java.exe to PATH, ",
        "then re-render."
      ))
    }
  }
}

library(rJava)
library(denguedatahub)
library(tidyverse)
library(plotly)
library(leaflet)
library(DT)
library(scales)
library(viridis)
library(lubridate)
library(tsibble)
library(zoo)
library(slider)
library(anomalize)
library(tibbletime)
```

```{r}
#| label: load-and-prepare-data
#| cache: true

# District-level weekly data
data_raw <- srilanka_weekly_data |>
  mutate(
    start.date = mdy(start.date),
    end.date   = mdy(end.date),
    district   = str_trim(district)
  ) |>
  filter(!is.na(cases), cases >= 0)

# National weekly totals
national_weekly <- data_raw |>
  group_by(year, week, start.date) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(start.date) |>
  mutate(yearweek = yearweek(start.date)) |>
  distinct(yearweek, .keep_all = TRUE)

# Annual totals
annual_totals <- data_raw |>
  group_by(year) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop")

# Value box stats
peak_year    <- annual_totals |> slice_max(total_cases, n = 1) |> pull(year)
peak_cases   <- annual_totals |> slice_max(total_cases, n = 1) |> pull(total_cases)
latest_year  <- max(data_raw$year)
latest_cases <- annual_totals |> filter(year == latest_year) |> pull(total_cases)
total_all    <- sum(annual_totals$total_cases)
n_districts  <- n_distinct(data_raw$district)

# Dynamic year range
year_min   <- min(data_raw$year)
year_max   <- max(data_raw$year)
year_range <- paste0(year_min, "-", year_max)

# Seasonality 
seasonality <- national_weekly |>
  group_by(week) |>
  summarise(
    avg_cases = mean(total_cases, na.rm = TRUE),
    min_cases = min(total_cases, na.rm = TRUE),
    max_cases = max(total_cases, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(roll_avg = zoo::rollmean(avg_cases, k = 3, fill = NA, align = "center"))

peak_week    <- seasonality |> slice_max(avg_cases, n = 1) |> pull(week)
peak_week_avg <- seasonality |> slice_max(avg_cases, n = 1) |> pull(avg_cases)

# District order by burden 
districts_ordered <- data_raw |>
  group_by(district) |>
  summarise(total = sum(cases, na.rm = TRUE)) |>
  arrange(desc(total)) |>
  pull(district)

#  Province mapping 
province_map <- tibble(
  district = c("Colombo","Gampaha","Kalutara",
               "Kandy","Matale","NuwaraEliya",
               "Galle","Matara","Hambantota",
               "Jaffna","Kilinochchi","Mannar","Vavuniya","Mullaitivu",
               "Batticaloa","Ampara","Trincomalee",
               "Kurunegala","Puttalam",
               "Anuradhapura","Polonnaruwa",
               "Badulla","Monaragala",
               "Ratnapura","Kegalle","Kalmune"),
  province = c("Western","Western","Western",
               "Central","Central","Central",
               "Southern","Southern","Southern",
               "Northern","Northern","Northern","Northern","Northern",
               "Eastern","Eastern","Eastern",
               "North Western","North Western",
               "North Central","North Central",
               "Uva","Uva",
               "Sabaragamuwa","Sabaragamuwa","Eastern")
)

district_province <- data_raw |>
  left_join(province_map, by = "district") |>
  group_by(year, province) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(province))

#  Epidemic threshold
threshold_df <- national_weekly |>
  group_by(week) |>
  summarise(threshold = mean(total_cases) + 2 * sd(total_cases), .groups = "drop")

national_threshold <- national_weekly |>
  left_join(threshold_df, by = "week") |>
  mutate(
    outbreak   = total_cases > threshold,
    rolling52  = slide_dbl(total_cases, mean, .before = 51, .complete = TRUE)
  )

#  Anomaly detection (anomalize / STL) 
# Convert to tbl_time for anomalize compatibility
anomaly_result <- national_weekly |>
  arrange(start.date) |>
  as_tbl_time(index = start.date) |>
  time_decompose(total_cases, method = "stl", frequency = 52,
                 message = FALSE) |>
  anomalize(remainder, method = "iqr") |>
  time_recompose()

# Separate anomalous points for plotting
anomaly_points <- anomaly_result |>
  filter(anomaly == "Yes")
```

# Overview

## Row {height="20%"}

### Column {width="25%"}

```{r}
#| content: valuebox
#| title: !expr paste0("Total Cases (", year_range, ")")
list(icon = "bug", color = "#c0392b", value = format(total_all, big.mark = ","))
```

### Column {width="25%"}

```{r}
#| content: valuebox
#| title: "Peak Outbreak Year"
list(icon = "exclamation-triangle", color = "#e67e22", value = as.character(peak_year))
```

### Column {width="25%"}

```{r}
#| content: valuebox
#| title: "Cases in Latest Year"
list(icon = "calendar-check", color = "#27ae60", value = format(latest_cases, big.mark = ","))
```

### Column {width="25%"}

```{r}
#| content: valuebox
#| title: "Districts Monitored"
list(icon = "map", color = "#2980b9", value = n_districts)
```

## Row {height="80%"}

### Column {width="25%"}

```{r}
#| title: "Weekly Dengue Cases (2006-2025)"
#| cache: true

# Compute Jan-1 tick positions for every year present in the data
years_in_data <- sort(unique(national_weekly$year))
jan_dates     <- as.Date(paste0(years_in_data, "-01-01"))

# 4-week rolling average, used as a toggle-able "smoothed" view
# (kept local to this chart so it doesn't alter national_weekly elsewhere)
weekly_plot_data <- national_weekly |>
  arrange(start.date) |>
  mutate(roll4 = zoo::rollmean(total_cases, k = 4, fill = NA, align = "right"))

plot_ly(weekly_plot_data, x = ~start.date) |>
  add_trace(
    y = ~total_cases,
    type = "scatter", mode = "lines",
    name = "Weekly cases",
    line = list(color = "#c0392b", width = 1.2),
    fill = "tozeroy", fillcolor = "rgba(192,57,43,0.10)",
    hovertemplate = "<b>%{x|%b %d, %Y}</b><br><b>%{y:,}</b> cases<extra></extra>"
  ) |>
  add_trace(
    y = ~roll4,
    type = "scatter", mode = "lines",
    name = "4-week average",
    visible = FALSE,
    line = list(color = "#8e44ad", width = 1.8),
    hovertemplate = "<b>%{x|%b %d, %Y}</b><br>4-wk avg: <b>%{y:,.0f}</b><extra></extra>"
  ) |>
  #  Marker for the very first data point (2006, week 52) 
  # so the start of the series is visible even though 2006 is a single week
  add_trace(
    data = weekly_plot_data |> filter(start.date == min(start.date)),
    x = ~start.date, y = ~total_cases,
    type = "scatter", mode = "markers",
    name = "Data start",
    marker = list(color = "#c0392b", size = 6, symbol = "circle-open", line = list(width = 2)),
    showlegend = FALSE,
    hovertemplate = "<b>Data begins: %{x|%b %d, %Y}</b><br><b>%{y:,}</b> cases<extra></extra>"
  ) |>
  layout(
    xaxis = list(
      title      = "Year",
      showgrid   = TRUE,
      gridcolor  = "rgba(0,0,0,0.06)",
      zeroline   = FALSE,
      # Force full data range so 2006-2007 always shows on initial load,
      # even if the range slider view gets reset
      range      = c(min(weekly_plot_data$start.date, na.rm = TRUE),
                      max(weekly_plot_data$start.date, na.rm = TRUE)),
      # One tick per year, placed at Jan 1, written horizontally along the axis
      tickmode   = "array",
      tickvals   = jan_dates,
      ticktext   = as.character(years_in_data),
      tickangle  = 0,
      ticks      = "outside",
      ticklen    = 4,
      tickfont   = list(size = 8),
      automargin = TRUE,
      # Thin vertical guide at every year boundary, so years are clearly
      # marked along the horizontal axis line
      rangeslider = list(visible = TRUE, thickness = 0.08),
      rangeselector = list(
        buttons = list(
          list(count = 1, label = "1Y", step = "year", stepmode = "backward"),
          list(count = 5, label = "5Y", step = "year", stepmode = "backward"),
          list(step = "all", label = "All")
        ),
        font = list(size = 8),
        x = 0, xanchor = "left",
        y = 1.42, yanchor = "top"
      )
    ),
    yaxis = list(
      title          = "Weekly Cases",
      showgrid       = TRUE,
      gridcolor      = "rgba(0,0,0,0.06)",
      zeroline       = FALSE,
      showticklabels = TRUE,
      tickformat     = ",d",
      tickfont       = list(size = 8),
      automargin     = TRUE
    ),
    shapes = lapply(jan_dates, function(d) {
      list(type = "line", x0 = d, x1 = d, y0 = 0, y1 = 1, yref = "paper",
           line = list(color = "rgba(0,0,0,0.08)", width = 1, dash = "dot"))
    }),
    annotations = list(
      list(
        x = 0.5, xref = "paper", xanchor = "center",
        y = 1.02, yref = "paper", yanchor = "bottom",
        text = "\u25CB Data begins Dec 2006 (week 52 only) \u2014 full weekly data from 2007 onward",
        showarrow = FALSE,
        font = list(size = 8, color = "#922b21"),
        bgcolor = "rgba(255,255,255,0.9)",
        bordercolor = "#c0392b", borderwidth = 1, borderpad = 3
      )
    ),
    updatemenus = list(
      list(
        type = "buttons", direction = "right",
        x = 0, xanchor = "left", y = 1.22, yanchor = "top",
        font = list(size = 8),
        buttons = list(
          list(method = "update",
               args = list(list(visible = c(TRUE, FALSE, TRUE))),
               label = "Weekly"),
          list(method = "update",
               args = list(list(visible = c(FALSE, TRUE, TRUE))),
               label = "4-wk avg")
        )
      )
    ),
    hovermode  = "x unified",
    margin     = list(t = 95, b = 60, l = 50, r = 5),
    showlegend = FALSE
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

### Column {width="25%"}

```{r}
#| title: "Annual Cases — Peak Year Highlighted"
#| cache: true

plot_ly(annual_totals, x = ~year, y = ~total_cases, type = "bar",
        marker = list(
          color = ifelse(annual_totals$year == peak_year, "#e67e22", "#f0d9b5"),
          line  = list(color = ifelse(annual_totals$year == peak_year, "#c0392b", "#d4a06a"),
                       width = 1.5),
          opacity = 0.92
        ),
        hovertemplate = "<b>%{x}</b><br>Cases: <b>%{y:,}</b><extra></extra>",
        selected    = list(marker = list(opacity = 1)),
        unselected  = list(marker = list(opacity = 0.4))) |>
  add_annotations(
    x = peak_year, y = peak_cases,
    text = paste0("<b>", peak_year, " ▲</b>"),
    showarrow = TRUE, arrowhead = 2, arrowsize = 0.8,
    ax = 0, ay = -30,
    font = list(color = "#e67e22", size = 10)
  ) |>
  layout(
    xaxis = list(title = "", showgrid = FALSE, tickangle = -45,
                 tickfont = list(size = 8), dtick = 1,
                 automargin = TRUE),
    yaxis = list(title = "", showgrid = TRUE,
                 gridcolor = "rgba(0,0,0,0.06)",
                 showticklabels = TRUE,
                 tickformat = ",d",
                 tickfont = list(size = 8),
                 automargin = TRUE),
    margin    = list(t = 10, b = 50, l = 55, r = 5),
    showlegend = FALSE,
    clickmode  = "event+select"
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

### Column {width="25%"}

```{r}
#| title: "Seasonal Pattern (Avg by Week)"
#| cache: true

latest_weekly <- national_weekly |>
  filter(year == latest_year) |>
  group_by(week) |>
  summarise(cases = sum(total_cases), .groups = "drop")

# Baseline for comparison against the current year: average of every OTHER
# year, excluding latest_year. 
seasonality_other_years <- national_weekly |>
  filter(year != latest_year) |>
  group_by(week) |>
  summarise(avg_cases = mean(total_cases, na.rm = TRUE), .groups = "drop")

plot_ly() |>
  add_trace(data = seasonality_other_years, x = ~week, y = ~avg_cases,
            type = "scatter", mode = "lines",
            name = "Avg of other years",
            line = list(color = "#aaaaaa", width = 1.5, dash = "dot"),
            fill = "tozeroy", fillcolor = "rgba(150,150,150,0.06)",
            hovertemplate = "Week %{x} avg (other years): <b>%{y:.0f}</b><extra></extra>") |>
  add_trace(data = latest_weekly, x = ~week, y = ~cases,
            type = "scatter", mode = "lines+markers",
            name = paste("Latest:", latest_year),
            line   = list(color = "#27ae60", width = 2),
            marker = list(color = "#27ae60", size = 4),
            hovertemplate = paste0("Week %{x} (", latest_year, "): <b>%{y:,}</b><extra></extra>")) |>
  layout(
    xaxis = list(title = "", showgrid = FALSE, zeroline = FALSE,
                 tickvals = c(1, 5, 9, 13, 18, 22, 26, 31, 35, 39, 44, 48),
                 ticktext = c("Jan","Feb","Mar","Apr","May","Jun",
                              "Jul","Aug","Sep","Oct","Nov","Dec"),
                 tickfont = list(size = 8),
                 tickangle = -45,
                 automargin = TRUE),
    yaxis = list(title = "Avg Cases", showgrid = TRUE,
                 gridcolor = "rgba(0,0,0,0.06)",
                 zeroline = FALSE,
                 showticklabels = TRUE,
                 tickformat = ",d",
                 tickfont = list(size = 8),
                 automargin = TRUE),
    shapes = lapply(c(1, 5, 9, 13, 18, 22, 26, 31, 35, 39, 44, 48), function(w) {
      list(type = "line",
           x0 = w, x1 = w,
           y0 = 0, y1 = 1,
           yref = "paper",
           line = list(color = "rgba(0,0,0,0.12)", width = 1, dash = "dot"))
    }),
    legend = list(orientation = "h", y = -0.28, font = list(size = 9),
                  bgcolor = "rgba(255,255,255,0)"),
    hovermode = "x unified",
    margin = list(t = 5, b = 55, l = 55, r = 5)
  )
```

### Column {width="25%"}

```{r}
#| title: "Cases by District (Map)"
#| cache: true

# Paired with: Districts Monitored — choropleth using lat/lon bubble map
district_coords <- tibble(
  district = c("Colombo","Gampaha","Kalutara","Kandy","Matale","NuwaraEliya",
               "Galle","Matara","Hambantota","Jaffna","Kilinochchi","Mannar",
               "Vavuniya","Mullaitivu","Batticaloa","Ampara","Trincomalee",
               "Kurunegala","Puttalam","Anuradhapura","Polonnaruwa",
               "Badulla","Monaragala","Ratnapura","Kegalle","Kalmune"),
  lat = c(6.93, 7.09, 6.58, 7.29, 7.47, 6.97,
          6.05, 5.95, 6.12, 9.66, 9.40, 8.98,
          8.75, 9.27, 7.72, 7.30, 8.57,
          7.47, 8.03, 8.33, 7.94,
          6.99, 6.87, 6.68, 7.25, 7.73),
  lon = c(79.85, 80.00, 80.22, 80.63, 80.62, 80.78,
          80.22, 80.54, 81.12, 80.02, 80.40, 79.90,
          80.50, 80.82, 81.70, 81.67, 81.23,
          80.36, 79.84, 80.41, 81.00,
          81.05, 81.35, 80.40, 80.35, 81.83)
)

district_map_data <- data_raw |>
  group_by(district) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  left_join(district_coords, by = "district") |>
  filter(!is.na(lat)) |>
  mutate(radius = scales::rescale(total_cases, to = c(6, 24)))

# Sky blue (low) → deep blue → muted purple → red → dark red (high)
pal <- colorNumeric(
  palette = c("#8ecae6", "#3a7ca5", "#8e6fa8", "#c0392b", "#7b0000"),
  domain  = district_map_data$total_cases
)

leaflet(district_map_data, options = leafletOptions(minZoom = 6, maxZoom = 11)) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius       = ~radius,
    fillColor    = ~pal(total_cases),
    fillOpacity  = 0.85,
    stroke       = TRUE, color = "#ffffff", weight = 1,
    label = ~paste0(district, ": ", format(total_cases, big.mark = ",")),
    popup = ~paste0(
      "<b>", district, "</b><br>",
      "Total cases: ", format(total_cases, big.mark = ","), " (", year_range, ")"
    ),
    labelOptions = labelOptions(
      style = list("font-weight" = "600", "font-size" = "12px"),
      textsize = "12px", direction = "auto"
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal, values = ~total_cases,
    title = "Cases", opacity = 0.9,
    labFormat = labelFormat(big.mark = ",")
  ) |>
  fitBounds(lng1 = 79.3, lat1 = 5.6, lng2 = 82.2, lat2 = 10.0) |>
 
  htmlwidgets::onRender("
    function(el, x) {
      var map = this;
      setTimeout(function() { map.invalidateSize(); }, 300);
      window.addEventListener('resize', function() { map.invalidateSize(); });
    }
  ")
```

# Seasonality & Trends

## Row {height="55%"}

### Column {width="55%"}

```{r}
#| title: "Weekly Dengue Seasonality — Average Cases & Historical Range"
#| cache: true

# Month-label positions (middle of each month in week numbers)
month_ticks <- c(2, 6, 10, 15, 19, 23, 28, 32, 36, 41, 45, 49)
month_labels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

y_ceil <- max(seasonality$max_cases, na.rm = TRUE) * 1.08

plot_ly(data = seasonality, x = ~week) |>

  #  Historical min/max ribbon 
  add_ribbons(
    ymin = ~min_cases,
    ymax = ~max_cases,
    name = "Historical range",
    fillcolor = "rgba(142,68,173,0.10)",
    line = list(color = "transparent"),
    hoverinfo = "skip",
    showlegend = TRUE
  ) |>

  #  Raw weekly average
  add_trace(
    data = seasonality,
    x = ~week,
    y = ~avg_cases,
    type = "scatter", mode = "lines+markers",
    name = "Avg all years",
    line   = list(color = "#5b2c6f", width = 1.8),
    marker = list(color = "#8e44ad", size = 5,
                  line = list(color = "white", width = 1.2)),
    customdata = ~paste0(round(min_cases), " – ", round(max_cases)),
    hovertemplate = paste0(
      "<b>Week %{x}</b><br>",
      "Average: <b>%{y:.0f}</b> cases<br>",
      "<i>Range: %{customdata}</i><extra></extra>"
    ),
    showlegend = TRUE
  ) |>

  #  Peak week annotation
  add_annotations(
    x = peak_week, y = peak_week_avg,
    text = paste0("<b>Peak: Wk ", peak_week, "<br>",
                  round(peak_week_avg), " avg cases</b>"),
    showarrow = TRUE, arrowhead = 2, arrowsize = 0.9,
    arrowcolor = "#5b2c6f",
    ax = 35, ay = -42,
    font = list(size = 10, color = "#5b2c6f"),
    bgcolor = "rgba(255,255,255,0.85)",
    bordercolor = "#8e44ad", borderpad = 4
  ) |>

  layout(
    xaxis = list(
      title      = "",
      showgrid   = TRUE,
      gridcolor  = "rgba(0,0,0,0.05)",
      zeroline   = FALSE,
      tickmode   = "array",
      tickvals   = month_ticks,
      ticktext   = month_labels,
      tickfont   = list(size = 11),
      automargin = TRUE
    ),
    yaxis = list(
      title     = "Weekly Cases (historical average)",
      showgrid  = TRUE,
      gridcolor = "rgba(0,0,0,0.05)",
      zeroline  = FALSE,
      tickfont  = list(size = 10)
    ),
    shapes = list(
      # SW Monsoon (May–Jul, weeks 18–30)
      list(type = "rect", x0 = 18, x1 = 30,
           y0 = 0, y1 = y_ceil,
           fillcolor = "rgba(255,140,0,0.09)",
           line = list(width = 0), layer = "below"),
      # NE Monsoon (Oct–Jan, weeks 41–52)
      list(type = "rect", x0 = 41, x1 = 52,
           y0 = 0, y1 = y_ceil,
           fillcolor = "rgba(30,100,220,0.08)",
           line = list(width = 0), layer = "below")
    ),
    annotations = list(
      list(
        x = 24, y = y_ceil * 0.97,
        text = "☂ SW Monsoon<br><i style='font-size:9px'>May–Jul</i>",
        showarrow = FALSE,
        font = list(size = 10, color = "darkorange"),
        align = "center"
      ),
      list(
        x = 47, y = y_ceil * 0.97,
        text = "☂ NE Monsoon<br><i style='font-size:9px'>Oct–Jan</i>",
        showarrow = FALSE,
        font = list(size = 10, color = "#1a5db5"),
        align = "center"
      )
    ),
    legend = list(
      orientation = "h", y = -0.18, x = 0,
      font = list(size = 10),
      bgcolor = "rgba(255,255,255,0)"
    ),
    hovermode  = "x unified",
    hoverlabel = list(bgcolor = "white", bordercolor = "#8e44ad",
                      font = list(size = 12)),
    margin = list(t = 20, b = 70, l = 10, r = 10)
  ) |>
  config(
    displayModeBar = TRUE,
    modeBarButtonsToRemove = c("lasso2d", "select2d"),
    displaylogo = FALSE,
    toImageButtonOptions = list(format = "png", filename = "seasonal_pattern")
  )
```

### Column {width="45%"}

```{r}
#| title: "Year-on-Year Comparison — Top 5 vs. All-Years Range"
#| cache: true

top5_years  <- annual_totals |> slice_max(total_cases, n = 5) |> pull(year)
weekly_top5 <- national_weekly |> filter(year %in% top5_years)
colors_yoy  <- c("#c0392b", "#e67e22", "#9b59b6", "#16a085", "#2980b9")

# Background ribbon: all-year min/max per week (already in `seasonality`)
p_yoy <- plot_ly() |>

  #  All-years min/max ribbon
  add_ribbons(
    data = seasonality,
    x = ~week,
    ymin = ~min_cases, ymax = ~max_cases,
    name = "All-years range",
    fillcolor = "rgba(32,178,170,0.15)",
    line = list(color = "rgba(32,178,170,0.35)", width = 0.8),
    hoverinfo = "skip",
    showlegend = TRUE
  )

#  Top-5 years
for (i in seq_along(top5_years)) {
  yr  <- top5_years[i]
  dat <- filter(weekly_top5, year == yr)
  p_yoy <- add_trace(
    p_yoy,
    x = dat$week, y = dat$total_cases,
    name = as.character(yr),
    type = "scatter", mode = "lines",
    line = list(color = colors_yoy[i], width = 2.2),
    hovertemplate = paste0(
      "<b>", yr, " — Week %{x}</b><br>",
      "<b>%{y:,}</b> cases<extra></extra>"
    )
  )
}

p_yoy |>
  layout(
    xaxis = list(
      title      = "",
      dtick      = 4,
      showgrid   = TRUE,
      gridcolor  = "rgba(0,0,0,0.05)",
      zeroline   = FALSE,
      tickmode   = "array",
      tickvals   = c(2, 6, 10, 15, 19, 23, 28, 32, 36, 41, 45, 49),
      ticktext   = c("Jan","Feb","Mar","Apr","May","Jun",
                     "Jul","Aug","Sep","Oct","Nov","Dec"),
      tickfont   = list(size = 10),
      automargin = TRUE,
      # Quick-zoom buttons
      rangeselector = list(
        buttons = list(
          list(label = "Q1", count = 13, step = "week", stepmode = "todate"),
          list(label = "Q2", count = 13, step = "week", stepmode = "todate"),
          list(label = "H1", count = 26, step = "week", stepmode = "backward"),
          list(label = "All", step = "all")
        ),
        font = list(size = 10),
        bgcolor = "#f4f4f4", activecolor = "#8e44ad"
      )
    ),
    yaxis = list(
      title     = "Weekly Cases",
      showgrid  = TRUE,
      gridcolor = "rgba(0,0,0,0.05)",
      zeroline  = FALSE,
      tickfont  = list(size = 10)
    ),
    legend = list(
      title       = list(text = "<b>Year</b>"),
      bgcolor     = "rgba(255,255,255,0.88)",
      bordercolor = "lightgrey", borderwidth = 1,
      font        = list(size = 10)
    ),
    hovermode  = "x unified",
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin = list(t = 50, b = 40, l = 10, r = 10)
  ) |>
  config(
    displayModeBar = TRUE,
    modeBarButtonsToRemove = c("lasso2d", "select2d"),
    displaylogo = FALSE,
    toImageButtonOptions = list(format = "png", filename = "yoy_comparison")
  )
```

## Row {height="45%"}

```{r}
#| title: "Seasonal Subseries — Monthly Case Distribution Across All Years"
#| cache: true

# for each calendar month,
# show the distribution of weekly cases across every year in the
# dataset, plus a connecting mean line 
subseries_data <- national_weekly |>
  mutate(month = lubridate::month(start.date, label = TRUE, abbr = TRUE)) |>
  filter(!is.na(month))

month_levels <- month.abb
subseries_data <- subseries_data |>
  mutate(month = factor(month, levels = month_levels))

month_means <- subseries_data |>
  group_by(month) |>
  summarise(mean_cases = mean(total_cases, na.rm = TRUE), .groups = "drop") |>
  arrange(match(month, month_levels))

plot_ly() |>
  add_trace(
    data = subseries_data,
    x = ~month, y = ~total_cases,
    type = "box",
    boxpoints = "outliers",
    marker  = list(color = "#5a9bd5", size = 3, opacity = 0.5),
    line    = list(color = "#2c5f8a", width = 1),
    fillcolor = "rgba(90,155,213,0.25)",
    hovertemplate = "<b>%{x}</b><br>Weekly cases: <b>%{y:,}</b><extra></extra>",
    name = "Weekly cases", showlegend = FALSE
  ) |>
  add_trace(
    data = month_means,
    x = ~month, y = ~mean_cases,
    type = "scatter", mode = "lines+markers",
    line   = list(color = "#c0392b", width = 2, dash = "dot"),
    marker = list(color = "#c0392b", size = 6, symbol = "diamond"),
    name = "Monthly mean",
    hovertemplate = "<b>%{x} mean</b><br><b>%{y:,.0f}</b> cases<extra></extra>"
  ) |>
  layout(
    xaxis = list(title = list(text = "\u25C6 Red diamonds = average weekly cases for that month, across all years",
                               font = list(size = 10.5, color = "#922b21")),
                 categoryorder = "array", categoryarray = month_levels,
                 showgrid = FALSE),
    yaxis = list(title = "Weekly cases", showgrid = TRUE,
                 gridcolor = "rgba(0,0,0,0.06)", zeroline = FALSE),
    legend = list(orientation = "h", x = 0, y = 1.1, font = list(size = 10)),
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin = list(t = 40, b = 55, l = 55, r = 10)
  ) |>
  config(
    displayModeBar = TRUE,
    modeBarButtonsToRemove = c("lasso2d", "select2d"),
    displaylogo = FALSE,
    toImageButtonOptions = list(format = "png", filename = "seasonal_subseries")
  )
```

# District & Province Analysis

## Row {height="50%"}

### Column {width="60%"}

```{r}
#| title: "Top 10 Districts by Dengue Cases — Select Year or All-Time"
#| cache: true

# Pre-compute per-year top-10 and all-time top-10
dist_by_year <- data_raw |>
  group_by(year, district) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop")

all_years_top10 <- dist_by_year |>
  group_by(district) |>
  summarise(total_cases = sum(total_cases), .groups = "drop") |>
  slice_max(total_cases, n = 10) |>
  arrange(total_cases)

available_years <- sort(unique(dist_by_year$year))

# Build one trace per year + one for all-time; toggle visibility via buttons
bar_colors <- function(vals) {
  rng <- range(vals)
  scales::col_numeric(
    palette = c("#fcbba1", "#d32f2f", "#7b0000"),
    domain  = rng
  )(vals)
}

p_bar <- plot_ly()

# Pre-sort each dataset so districts appear low→high 
all_years_top10 <- all_years_top10 |> arrange(total_cases)

# Store per-year sorted data for button args
yr_datasets <- lapply(available_years, function(yr) {
  dist_by_year |>
    filter(year == yr) |>
    slice_max(total_cases, n = 10) |>
    arrange(total_cases)          # ascending so highest is at top of h-bar
})

# All-time trace (visible by default)
p_bar <- add_trace(
  p_bar,
  x       = all_years_top10$total_cases,
  y       = all_years_top10$district,   
  type    = "bar", orientation = "h",
  name    = "All Years",
  visible = TRUE,
  marker  = list(
    color = bar_colors(all_years_top10$total_cases),
    line  = list(color = "rgba(0,0,0,0.08)", width = 0.5)
  ),
  text         = format(all_years_top10$total_cases, big.mark = ","),
  textposition = "outside",
  textfont     = list(size = 9),
  hovertemplate = "<b>%{y}</b><br>Total cases: <b>%{x:,}</b><extra></extra>"
)

# Per-year traces (hidden initially)
for (i in seq_along(available_years)) {
  yr     <- available_years[i]
  yr_data <- yr_datasets[[i]]
  p_bar <- add_trace(
    p_bar,
    x       = yr_data$total_cases,
    y       = yr_data$district,          # plain character per year
    type    = "bar", orientation = "h",
    name    = as.character(yr),
    visible = FALSE,
    marker  = list(
      color = bar_colors(yr_data$total_cases),
      line  = list(color = "rgba(0,0,0,0.08)", width = 0.5)
    ),
    text         = format(yr_data$total_cases, big.mark = ","),
    textposition = "outside",
    textfont     = list(size = 9),
    hovertemplate = paste0("<b>%{y}</b><br>", yr, " cases: <b>%{x:,}</b><extra></extra>")
  )
}

n_traces <- 1 + length(available_years)
make_vis <- function(i) { v <- rep(FALSE, n_traces); v[i] <- TRUE; v }

# Dropdown buttons — each button also updates the y-axis categoryarray so
# districts re-rank correctly for that selection
btn_list <- c(
  list(list(
    method = "update",
    args   = list(
      list(visible = make_vis(1)),
      list(yaxis = list(
        categoryorder = "array",
        categoryarray = all_years_top10$district,   # low→high → top bar = highest
        automargin    = TRUE,
        tickfont      = list(size = 11)
      ))
    ),
    label = "All Years"
  )),
  lapply(seq_along(available_years), function(i) {
    yr_data <- yr_datasets[[i]]
    list(
      method = "update",
      args   = list(
        list(visible = make_vis(i + 1)),
        list(yaxis = list(
          categoryorder = "array",
          categoryarray = yr_data$district,          # re-ranked for this year
          automargin    = TRUE,
          tickfont      = list(size = 11)
        ))
      ),
      label = as.character(available_years[i])
    )
  })
)

p_bar |>
  layout(
    xaxis    = list(title = "Total Cases", showgrid = TRUE,
                    gridcolor = "rgba(0,0,0,0.06)", automargin = TRUE),
    yaxis    = list(title = "", automargin = TRUE, tickfont = list(size = 11)),
    showlegend = FALSE,
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin   = list(t = 50, l = 110, r = 70, b = 40),
    updatemenus = list(list(
      type       = "dropdown",
      direction  = "down",
      x          = 0.0, xanchor = "left",
      y          = 1.18, yanchor = "top",
      bgcolor    = "#f8f8f8",
      bordercolor = "#cccccc",
      font       = list(size = 11),
      buttons    = btn_list
    ))
  ) |>
  config(
    displayModeBar = TRUE,
    modeBarButtonsToRemove = c("lasso2d", "select2d"),
    displaylogo = FALSE,
    toImageButtonOptions = list(format = "png", filename = "top10_districts")
  )
```

### Column {width="40%"}

```{r}
#| title: "Province-Level Annual Trends — Linear & Log Scale"
#| cache: true

province_colors <- c(
  "Western"       = "#c0392b",
  "Central"       = "#e67e22",
  "Southern"      = "#f1c40f",
  "Northern"      = "#27ae60",
  "Eastern"       = "#2980b9",
  "North Western" = "#8e44ad",
  "North Central" = "#16a085",
  "Uva"           = "#d35400",
  "Sabaragamuwa"  = "#2c3e50"
)

provinces <- unique(district_province$province)

p_prov <- plot_ly()
for (prov in provinces) {
  pdat <- filter(district_province, province == prov) |> arrange(year)
  p_prov <- add_trace(
    p_prov,
    x    = pdat$year, y = pdat$total_cases,
    type = "scatter", mode = "lines+markers",
    name = prov,
    line   = list(color = province_colors[prov], width = 2),
    marker = list(color = province_colors[prov], size = 6,
                  line = list(color = "white", width = 1)),
    hovertemplate = paste0(
      "<b>", prov, "</b><br>",
      "Year: %{x}<br>",
      "Cases: <b>%{y:,}</b><extra></extra>"
    )
  )
}

p_prov |>
  layout(
    xaxis = list(
      title     = "Year", dtick = 2, tickangle = -45,
      showgrid  = TRUE, gridcolor = "rgba(0,0,0,0.05)",
      tickfont  = list(size = 10)
    ),
    yaxis = list(
      title     = "Total Cases",
      showgrid  = TRUE, gridcolor = "rgba(0,0,0,0.05)",
      tickfont  = list(size = 10)
    ),
    hovermode  = "x unified",
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    legend = list(
      title       = list(text = "<b>Province</b>"),
      bgcolor     = "rgba(255,255,255,0.88)",
      bordercolor = "lightgrey", borderwidth = 1,
      font        = list(size = 10)
    ),
    margin = list(t = 55, b = 50, l = 10, r = 10),
    # Log / Linear toggle
    updatemenus = list(list(
      type       = "buttons",
      direction  = "right",
      x          = 0.0, xanchor = "left",
      y          = 1.18, yanchor = "top",
      bgcolor    = "#f8f8f8",
      bordercolor = "#cccccc",
      font       = list(size = 11),
      buttons    = list(
        list(method = "relayout", args = list(list(yaxis.type = "linear")),
             label = "Linear"),
        list(method = "relayout", args = list(list(yaxis.type = "log")),
             label = "Log scale")
      )
    ))
  ) |>
  config(
    displayModeBar = TRUE,
    modeBarButtonsToRemove = c("lasso2d", "select2d"),
    displaylogo = FALSE,
    toImageButtonOptions = list(format = "png", filename = "province_trends")
  )
```

## Row {height="50%"}

```{r}
#| title: "Weekly Dengue Intensity by Province — Browse by Year"
#| cache: true

heat_prov_full <- data_raw |>
  left_join(province_map, by = "district") |>
  filter(!is.na(province), year > 2006, week >= 1, week <= 52) |>
  group_by(year, week, province) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

# Province order: lowest total burden first 
province_order <- heat_prov_full |>
  group_by(province) |>
  summarise(total = sum(cases), .groups = "drop") |>
  arrange(total) |>
  pull(province)

prov_yrs  <- sort(unique(heat_prov_full$year))
prov_zmax <- quantile(heat_prov_full$cases, 0.99, na.rm = TRUE)

prov_month_tick_vals <- c(1, 5, 9, 14, 18, 22, 27, 31, 35, 40, 44, 48)
prov_month_tick_text <- c("Jan","Feb","Mar","Apr","May","Jun",
                          "Jul","Aug","Sep","Oct","Nov","Dec")

# All (week, province) combinations for a complete grid
prov_grid <- tidyr::expand_grid(
  province = factor(province_order, levels = province_order),
  week     = 1:52
)

p_heat_prov <- plot_ly()

for (i in seq_along(prov_yrs)) {
  yr <- prov_yrs[i]

  # Join actual data onto the full grid so every cell is present
  ydat <- prov_grid |>
    left_join(
      heat_prov_full |> filter(year == yr) |> select(province, week, cases),
      by = c("province", "week")
    ) |>
    mutate(
      cases    = replace_na(cases, 0),
      province = factor(province, levels = province_order)
    ) |>
    arrange(province, week)   # province_order row order, weeks 1-52 column order

  p_heat_prov <- add_trace(
    p_heat_prov,
    x    = ydat$week,
    y    = as.character(ydat$province),
    z    = ydat$cases,
    type = "heatmap",
    visible = (i == 1),
    colorscale = list(
      list(0,    "#f0f4ff"), list(0.15, "#9ecae1"), list(0.35, "#9b59b6"),
      list(0.55, "#e67e22"), list(0.75, "#e74c3c"), list(0.88, "#c0392b"), list(1,    "#641e16")
    ),
    zmin = 0, zmax = prov_zmax,
    colorbar = list(title = "<b>Cases/<br>week</b>", thickness = 14,
                    tickfont = list(size = 9), len = 0.85, outlinewidth = 0),
    hovertemplate = paste0("<b>%{y}</b><br>Week %{x} (", yr,
                           ")<br>Cases: <b>%{z:,}</b><extra></extra>"),
    xgap = 1, ygap = 0.6, showscale = (i == 1)
  )
}

prov_year_buttons <- lapply(seq_along(prov_yrs), function(i) {
  vis <- rep(FALSE, length(prov_yrs)); vis[i] <- TRUE
  list(method = "update", args = list(list(visible = vis, showscale = vis)),
       label = as.character(prov_yrs[i]))
})

p_heat_prov |>
  layout(
    xaxis = list(
      title    = "", range = list(0.5, 52.5),
      tickmode = "array", tickvals = prov_month_tick_vals,
      ticktext = prov_month_tick_text, tickfont = list(size = 10),
      showgrid = FALSE, automargin = TRUE
    ),
    yaxis = list(
      title         = "",
      categoryorder = "array",
      categoryarray = province_order,
      automargin    = TRUE, showgrid = FALSE, tickfont = list(size = 10)
    ),
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin = list(t = 55, b = 30, l = 120, r = 20),
    updatemenus = list(list(
      type = "dropdown", direction = "down",
      x = 0.0, xanchor = "left", y = 1.08, yanchor = "top",
      bgcolor = "#f8f8f8", bordercolor = "#cccccc",
      font = list(size = 11), buttons = prov_year_buttons
    ))
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d", "select2d"),
         displaylogo = FALSE,
         toImageButtonOptions = list(format = "png", filename = "province_week_heatmap"))
```

## Row {height="50%"}

```{r}
#| title: "Weekly Dengue Intensity by District — Browse by Year"
#| cache: true

heat_dist_full <- data_raw |>
  filter(year > 2006, week >= 1, week <= 52) |>
  group_by(year, week, district) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

# districts_ordered is highest-burden first
dist_order <- rev(districts_ordered)

all_yrs   <- sort(unique(heat_dist_full$year))
dist_zmax <- quantile(heat_dist_full$cases, 0.99, na.rm = TRUE)

month_tick_vals <- c(1, 5, 9, 14, 18, 22, 27, 31, 35, 40, 44, 48)
month_tick_text <- c("Jan","Feb","Mar","Apr","May","Jun",
                     "Jul","Aug","Sep","Oct","Nov","Dec")

# All (week, district) combinations for a complete grid
dist_grid <- tidyr::expand_grid(
  district = factor(dist_order, levels = dist_order),
  week     = 1:52
)

p_heat <- plot_ly()

for (i in seq_along(all_yrs)) {
  yr <- all_yrs[i]

  ydat <- dist_grid |>
    left_join(
      heat_dist_full |> filter(year == yr) |> select(district, week, cases),
      by = c("district", "week")
    ) |>
    mutate(
      cases    = replace_na(cases, 0),
      district = factor(district, levels = dist_order)
    ) |>
    arrange(district, week)   # dist_order row order, weeks 1-52 column order

  p_heat <- add_trace(
    p_heat,
    x    = ydat$week,
    y    = as.character(ydat$district),
    z    = ydat$cases,
    type = "heatmap",
    visible = (i == 1),
    colorscale = list(
      list(0,    "#f0f4ff"), list(0.15, "#9ecae1"), list(0.35, "#9b59b6"),
      list(0.55, "#e67e22"), list(0.75, "#e74c3c"), list(0.88, "#c0392b"), list(1,    "#641e16")
    ),
    zmin = 0, zmax = dist_zmax,
    colorbar = list(title = "<b>Cases/<br>week</b>", thickness = 14,
                    tickfont = list(size = 9), len = 0.85, outlinewidth = 0),
    hovertemplate = paste0("<b>%{y}</b><br>Week %{x} (", yr,
                           ")<br>Cases: <b>%{z:,}</b><extra></extra>"),
    xgap = 1, ygap = 0.6, showscale = (i == 1)
  )
}

year_buttons <- lapply(seq_along(all_yrs), function(i) {
  vis <- rep(FALSE, length(all_yrs)); vis[i] <- TRUE
  list(method = "update", args = list(list(visible = vis, showscale = vis)),
       label = as.character(all_yrs[i]))
})

p_heat |>
  layout(
    xaxis = list(
      title    = "", range = list(0.5, 52.5),
      tickmode = "array", tickvals = month_tick_vals,
      ticktext = month_tick_text, tickfont = list(size = 10),
      showgrid = FALSE, automargin = TRUE
    ),
    yaxis = list(
      title         = "",
      categoryorder = "array",
      categoryarray = dist_order,
      automargin    = TRUE, showgrid = FALSE, tickfont = list(size = 9)
    ),
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    margin = list(t = 55, b = 30, l = 120, r = 20),
    updatemenus = list(list(
      type = "dropdown", direction = "down",
      x = 0.0, xanchor = "left", y = 1.08, yanchor = "top",
      bgcolor = "#f8f8f8", bordercolor = "#cccccc",
      font = list(size = 11), buttons = year_buttons
    ))
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d", "select2d"),
         displaylogo = FALSE,
         toImageButtonOptions = list(format = "png", filename = "district_week_heatmap"))
```

# Outbreak Analysis

## Row {height="40%"}

```{r}
#| title: "Epidemic Channel: Weekly Cases vs. Threshold (mean + 2 SD) with 52-Week Rolling Average"
#| cache: true

#  Compute mean ribbon (mean alone, and mean + 2SD) per week
ribbon_df <- national_weekly |>
  group_by(week) |>
  summarise(
    mean_cases = mean(total_cases, na.rm = TRUE),
    sd_cases   = sd(total_cases,   na.rm = TRUE),
    .groups    = "drop"
  ) |>
  mutate(
    ribbon_lo = pmax(mean_cases - sd_cases, 0),
    ribbon_hi = mean_cases + 2 * sd_cases
  )

# Join ribbon bounds + rolling mean back onto the weekly time series
nat_plot <- national_threshold |>
  left_join(ribbon_df |> select(week, ribbon_lo, ribbon_hi, mean_cases),
            by = "week")

#  Major outbreak year annotations
outbreak_annotations <- list(
  list(x = as.Date("2017-06-01"), text = "2017<br>epidemic", color = "#c0392b"),
  list(x = as.Date("2019-07-01"), text = "2019<br>outbreak", color = "#e67e22")
)

y_max <- max(nat_plot$total_cases, na.rm = TRUE)

p_channel <- plot_ly(nat_plot, x = ~start.date)

#  1. Shaded ribbon: mean ± 1SD → mean + 2SD (the "alert zone") 
p_channel <- p_channel |>
  add_ribbons(
    ymin = ~ribbon_lo, ymax = ~ribbon_hi,
    name      = "Expected range (mean ± 1–2 SD)",
    fillcolor = "rgba(180,180,180,0.22)",
    line      = list(color = "transparent"),
    hoverinfo = "skip",
    showlegend = TRUE
  )

#  2. Threshold line (mean + 2SD) 
p_channel <- p_channel |>
  add_lines(
    y    = ~ribbon_hi,
    name = "Epidemic threshold (mean + 2 SD)",
    line = list(color = "rgba(192,57,43,0.50)", width = 1.2, dash = "dot"),
    hovertemplate = "<b>Threshold %{x|%b %Y}</b>: %{y:,.0f}<extra></extra>"
  )

#  3. Filled segments only where above threshold (epidemic periods)
above_thresh <- nat_plot |> filter(outbreak)
p_channel <- p_channel |>
  add_trace(
    data = above_thresh,
    x = ~start.date, y = ~total_cases,
    type = "scatter", mode = "none",
    fill = "tozeroy",
    fillcolor = "rgba(192,57,43,0.30)",
    name = "Epidemic period (above threshold)",
    hovertemplate = "<b>⚠ %{x|%d %b %Y}</b><br>Cases: <b>%{y:,}</b><extra></extra>"
  )

#  4. Weekly cases line
p_channel <- p_channel |>
  add_lines(
    y    = ~total_cases,
    name = "Weekly Cases",
    line = list(color = "#2471a3", width = 1.8),
    hovertemplate = "<b>%{x|%d %b %Y}</b><br>Cases: <b>%{y:,}</b><extra></extra>"
  )

#  5. 52-week rolling average
p_channel <- p_channel |>
  add_lines(
    y    = ~rolling52,
    name = "52-week rolling average",
    line = list(color = "#27ae60", width = 2, dash = "dash"),
    hovertemplate = "<b>%{x|%b %Y}</b><br>52-wk avg: <b>%{y:,.0f}</b><extra></extra>"
  )

#  6. Outbreak year annotations
for (ann in outbreak_annotations) {
  p_channel <- p_channel |>
    add_annotations(
      x         = ann$x,
      y         = y_max * 0.98,
      text      = paste0("<b>", ann$text, "</b>"),
      showarrow = FALSE,
      font      = list(size = 10, color = ann$color),
      bgcolor   = "rgba(255,255,255,0.82)",
      bordercolor = ann$color,
      borderpad = 3
    )
}

p_channel |>
  layout(
    xaxis = list(
      title    = "Date",
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)",
      rangeselector = list(
        buttons = list(
          list(count = 2,  label = "2yr",  step = "year",  stepmode = "backward"),
          list(count = 5,  label = "5yr",  step = "year",  stepmode = "backward"),
          list(count = 10, label = "10yr", step = "year",  stepmode = "backward"),
          list(step = "all", label = "All")
        )
      ),
      rangeslider = list(visible = TRUE, thickness = 0.05)
    ),
    yaxis = list(
      title    = "Weekly Cases",
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)",
      rangemode = "tozero"
    ),
    legend = list(
      orientation = "h", y = -0.30,
      bgcolor = "rgba(255,255,255,0.85)",
      font    = list(size = 10)
    ),
    hovermode = "x unified",
    margin    = list(t = 30, b = 100)
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

## Row {height="40%"}

```{r}
#| title: "Anomaly Detection — STL Decomposition + IQR Flagging (anomalize)"
#| cache: true

#  Observed series with anomalies highlighted
normal_pts  <- anomaly_result |> filter(anomaly == "No")
anomaly_pts <- anomaly_result |> filter(anomaly == "Yes")

p_anom <- plot_ly()

# Expected band (recomposed bounds)
p_anom <- p_anom |>
  add_ribbons(
    data      = anomaly_result,
    x         = ~start.date,
    ymin      = ~recomposed_l1,
    ymax      = ~recomposed_l2,
    name      = "Expected band (IQR)",
    fillcolor = "rgba(52,152,219,0.12)",
    line      = list(color = "transparent"),
    hoverinfo = "skip",
    showlegend = TRUE
  )

# Observed line
p_anom <- p_anom |>
  add_lines(
    data = anomaly_result,
    x    = ~start.date, y = ~observed,
    name = "Weekly Cases (observed)",
    line = list(color = "rgba(44,62,80,0.55)", width = 1.2),
    hovertemplate = "<b>%{x|%d %b %Y}</b><br>Cases: <b>%{y:,}</b><extra></extra>"
  )

# Normal points (small, muted)
p_anom <- p_anom |>
  add_markers(
    data   = normal_pts,
    x      = ~start.date, y = ~observed,
    name   = "Normal week",
    marker = list(color = "rgba(52,152,219,0.55)", size = 4),
    hovertemplate = "%{x|%d %b %Y}: <b>%{y:,}</b><extra></extra>"
  )

# Anomalous points (prominent red)
p_anom <- p_anom |>
  add_markers(
    data   = anomaly_pts,
    x      = ~start.date, y = ~observed,
    name   = "Anomalous week ⚠",
    marker = list(color = "#c0392b", size = 7,
                  line = list(color = "white", width = 1.2)),
    hovertemplate = "<b>⚠ Anomaly: %{x|%d %b %Y}</b><br>Cases: <b>%{y:,}</b><extra></extra>"
  )

p_anom |>
  layout(
    xaxis = list(
      title    = "Date",
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)",
      rangeselector = list(
        buttons = list(
          list(count = 2,  label = "2yr",  step = "year",  stepmode = "backward"),
          list(count = 5,  label = "5yr",  step = "year",  stepmode = "backward"),
          list(count = 10, label = "10yr", step = "year",  stepmode = "backward"),
          list(step = "all", label = "All")
        )
      ),
      rangeslider = list(visible = TRUE, thickness = 0.05)
    ),
    yaxis = list(
      title    = "Weekly Cases",
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)",
      rangemode = "tozero"
    ),
    legend = list(
      orientation = "h", y = -0.30,
      bgcolor     = "rgba(255,255,255,0.85)",
      font        = list(size = 10)
    ),
    hovermode = "x unified",
    margin    = list(t = 30, b = 100)
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

## Row {height="20%"}

### Column {width="50%"}

```{r}
#| title: "Cumulative Cases by Year"
#| cache: true

cum_data <- national_weekly |>
  group_by(year) |> arrange(week) |>
  mutate(cum_cases = cumsum(total_cases)) |> ungroup()

highlight_years <- c(2017, 2019, 2023, 2024, 2025)
other_years     <- setdiff(unique(cum_data$year), highlight_years)
hi_colors       <- c("#c0392b","#e67e22","#8e44ad","#2980b9","#27ae60")

#  Median & IQR ribbon across all non-highlighted years per week
ribbon_cum <- cum_data |>
  filter(year %in% other_years) |>
  group_by(week) |>
  summarise(
    med_cum = median(cum_cases, na.rm = TRUE),
    lo_cum  = quantile(cum_cases, 0.25, na.rm = TRUE),
    hi_cum  = quantile(cum_cases, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

p_cum <- plot_ly()

#  1. IQR ribbon for "typical" years
p_cum <- p_cum |>
  add_ribbons(
    data      = ribbon_cum,
    x         = ~week,
    ymin      = ~lo_cum,
    ymax      = ~hi_cum,
    name      = "Typical range (IQR of other years)",
    fillcolor = "rgba(150,150,150,0.18)",
    line      = list(color = "transparent"),
    hoverinfo = "skip",
    showlegend = TRUE
  ) |>
  # Median line
  add_lines(
    data = ribbon_cum,
    x    = ~week, y = ~med_cum,
    name = "Median (other years)",
    line = list(color = "rgba(130,130,130,0.75)", width = 1.4, dash = "dot"),
    hovertemplate = "Median Wk %{x}: <b>%{y:,}</b><extra></extra>"
  )

#  2. Highlighted years with direct end-labels
for (i in seq_along(highlight_years)) {
  yr <- highlight_years[i]
  d  <- filter(cum_data, year == yr)
  if (nrow(d) == 0) next

  last_pt <- d |> filter(week == max(week))

  p_cum <- p_cum |>
    add_trace(
      x    = d$week, y = d$cum_cases,
      type = "scatter", mode = "lines",
      name = as.character(yr),
      line = list(color = hi_colors[i], width = 2.5),
      hovertemplate = paste0("<b>", yr, " — Week %{x}</b><br>",
                             "Cumulative: <b>%{y:,}</b><extra></extra>")
    ) |>
    add_annotations(
      x         = last_pt$week,
      y         = last_pt$cum_cases,
      text      = paste0("<b>", yr, "</b>"),
      xanchor   = "left",
      showarrow = FALSE,
      font      = list(size = 10, color = hi_colors[i]),
      xshift    = 4
    )
}

#  3. Week-26 midyear reference line
p_cum |>
  layout(
    xaxis = list(
      title    = "Epidemiological Week",
      dtick    = 4,
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)"
    ),
    yaxis = list(
      title    = "Cumulative Cases",
      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)"
    ),
    shapes = list(
      list(
        type = "line", x0 = 26, x1 = 26, y0 = 0, y1 = 1, yref = "paper",
        line = list(color = "rgba(100,100,100,0.35)", width = 1.2, dash = "dot")
      )
    ),
    annotations = list(
      list(
        x = 26, y = 1, yref = "paper",
        text = "Mid-year (Wk 26)", showarrow = FALSE,
        font = list(size = 9, color = "grey50"),
        xanchor = "left", yanchor = "top", xshift = 4
      )
    ),
    legend = list(
      title       = list(text = "<b>Year</b>"),
      bgcolor     = "rgba(255,255,255,0.85)",
      bordercolor = "lightgrey", borderwidth = 1,
      font        = list(size = 10)
    ),
    hovermode = "closest",
    margin    = list(t = 30, r = 60)
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

### Column {width="50%"}

```{r}
#| title: "5-Year Rolling Average vs. Annual Cases"
#| cache: true

rolling <- annual_totals |>
  arrange(year) |>
  mutate(rolling5 = zoo::rollmean(total_cases, k = 5, fill = NA, align = "right"))

#  Color bars: red if above rolling avg, steel-blue if below
rolling <- rolling |>
  mutate(
    bar_color     = case_when(
      is.na(rolling5)              ~ "rgba(150,150,150,0.45)",   # no avg yet
      total_cases >= rolling5      ~ "rgba(192,57,43,0.60)",     # above avg → red
      TRUE                         ~ "rgba(52,152,219,0.50)"     # below avg → blue
    ),
    bar_color_bdr = case_when(
      is.na(rolling5)              ~ "rgba(120,120,120,0.8)",
      total_cases >= rolling5      ~ "rgba(152,41,31,0.9)",
      TRUE                         ~ "rgba(31,120,180,0.9)"
    )
  )

#  Reference value: rolling avg at the most recent complete year
last_rolling_val  <- rolling |> filter(!is.na(rolling5)) |> slice_max(year) |>
  pull(rolling5)
last_rolling_year <- rolling |> filter(!is.na(rolling5)) |> slice_max(year) |>
  pull(year)

plot_ly(rolling, x = ~year) |>

  #  Bars colored by above/below status
  add_bars(
    y      = ~total_cases,
    name   = "Annual Cases",
    marker = list(
      color = ~bar_color,
      line  = list(color = ~bar_color_bdr, width = 1.2)
    ),
    hovertemplate = paste0(
      "<b>%{x}</b><br>",
      "Annual cases: <b>%{y:,}</b><extra></extra>"
    )
  ) |>

  #  Rolling average line
  add_lines(
    y    = ~rolling5,
    name = "5-yr rolling avg",
    line = list(color = "#2c3e50", width = 2.5),
    hovertemplate = "<b>%{x}</b><br>5-yr avg: <b>%{y:,.0f}</b><extra></extra>"
  ) |>

  #  Peak year annotation
  add_annotations(
    x         = rolling |> slice_max(total_cases) |> pull(year),
    y         = rolling |> slice_max(total_cases) |> pull(total_cases),
    text      = paste0("<b>",
                       format(rolling |> slice_max(total_cases) |> pull(total_cases),
                              big.mark = ","),
                       " cases</b>"),
    showarrow = TRUE, arrowhead = 2, arrowsize = 0.8,
    ax = 0, ay = -36,
    font      = list(size = 10, color = "#922b21"),
    bgcolor   = "rgba(255,255,255,0.85)",
    bordercolor = "#c0392b", borderpad = 3
  ) |>

  layout(
    xaxis = list(
      title     = "Year",
      dtick     = 1, tickangle = -45,
      showgrid  = FALSE, tickfont = list(size = 9)
    ),
    yaxis = list(
      title     = "Cases",
      showgrid  = TRUE, gridcolor = "rgba(0,0,0,0.06)",
      rangemode = "tozero"
    ),
    # Horizontal reference line at most recent rolling avg
    shapes = list(
      list(
        type = "line",
        x0 = min(rolling$year), x1 = max(rolling$year),
        y0 = last_rolling_val,  y1 = last_rolling_val,
        line = list(color = "rgba(44,62,80,0.30)", width = 1.2, dash = "dot")
      )
    ),
    annotations = list(
      list(
        x       = max(rolling$year),
        y       = last_rolling_val,
        text    = paste0("<i>", last_rolling_year, " avg: ",
                         format(round(last_rolling_val), big.mark = ","), "</i>"),
        xanchor = "right", yanchor = "bottom",
        showarrow = FALSE,
        font    = list(size = 9, color = "rgba(44,62,80,0.70)")
      ),
      # Legend for bar colors
      list(
        x = min(rolling$year) + 1, y = max(rolling$total_cases, na.rm=TRUE) * 0.96,
        text = "<span style='color:#c0392b'>■</span> Above 5-yr avg  " ,
        showarrow = FALSE, font = list(size = 10), xanchor = "left"
      ),
      list(
        x = min(rolling$year) + 8, y = max(rolling$total_cases, na.rm=TRUE) * 0.96,
        text = "<span style='color:#2980b9'>■</span> Below 5-yr avg",
        showarrow = FALSE, font = list(size = 10), xanchor = "left"
      )
    ),
    hovermode  = "x unified",
    barmode    = "overlay",
    showlegend = TRUE,
    legend = list(
      orientation = "h", y = -0.30,
      bgcolor = "rgba(255,255,255,0.85)",
      font    = list(size = 10)
    ),
    margin = list(t = 30, b = 80)
  ) |>
  config(displayModeBar = TRUE, modeBarButtonsToRemove = c("lasso2d","select2d"),
         displaylogo = FALSE)
```

# Data Explorer

## Row

```{r}
#| title: "District-Level Weekly Dataset"

date_filter_ui <- htmltools::tags$div(
  style = "display:flex; align-items:center; gap:16px; margin-bottom:8px; flex-wrap:wrap;",
  htmltools::tags$div(
    style = "display:flex; align-items:center; gap:8px;",
    htmltools::tags$label(
      "Start date:", style = "font-weight:600; margin:0;", `for` = "explorerStartDate"
    ),
    htmltools::tags$input(
      type = "text", id = "explorerStartDate",
      placeholder = "Select start date",
      class = "form-control form-control-sm",
      style = "max-width:160px;"
    )
  ),
  htmltools::tags$div(
    style = "display:flex; align-items:center; gap:8px;",
    htmltools::tags$label(
      "End date:", style = "font-weight:600; margin:0;", `for` = "explorerEndDate"
    ),
    htmltools::tags$input(
      type = "text", id = "explorerEndDate",
      placeholder = "Select end date",
      class = "form-control form-control-sm",
      style = "max-width:160px;"
    )
  ),
  htmltools::tags$div(
    style = "display:flex; align-items:center; gap:8px;",
    htmltools::tags$label(
      "Cases:", style = "font-weight:600; margin:0;", `for` = "explorerCasesMin"
    ),
    htmltools::tags$input(
      type = "number", id = "explorerCasesMin", step = "10", placeholder = "Min",
      class = "form-control form-control-sm", style = "max-width:90px;"
    ),
    htmltools::tags$span("to"),
    htmltools::tags$input(
      type = "number", id = "explorerCasesMax", step = "10", placeholder = "Max",
      class = "form-control form-control-sm", style = "max-width:90px;"
    )
  ),
  htmltools::tags$button(
    "\u21bb Refresh",
    id    = "explorerRefreshBtn",
    type  = "button",
    class = "btn btn-sm btn-outline-secondary",
    style = "margin-left:auto;"
  )
)

explorer_table <- data_raw |>
  select(Date = start.date, Week = week, District = district, Cases = cases) |>
  arrange(Date, District) |>
  mutate(
    Week     = factor(Week,     levels = sort(unique(Week))),
    District = factor(District, levels = sort(unique(District))),
    Cases    = as.integer(Cases)
  ) |>
  datatable(
    elementId = "explorerTable",
    filter    = "top",
    rownames  = FALSE,
    options   = list(
      pageLength     = 15,
      lengthMenu     = c(15, 25, 50, 100),
      scrollX        = TRUE,
      deferRender    = TRUE,     # render rows only when needed
      paging         = TRUE,     
      dom            = "Blfrtip",
      buttons        = c("csv", "excel"),
      columnDefs  = list(
        list(className  = "dt-center", targets = c(0,1,2,3)),
        list(searchable = FALSE,       targets = c(0,3))
      ),                                                    
      initComplete = JS("
        function(settings) {
          var api = this.api();
          var tableId = api.table().node().id;

          var state = { startDate: null, endDate: null, casesMin: null, casesMax: null };

          // Row dates come through as 'YYYY-MM-DD' strings. new Date('YYYY-MM-DD')
          // parses as UTC midnight, while flatpickr's selected dates are LOCAL
          // midnight — in timezones ahead of UTC that mismatch silently drops the
          // selected end date's rows. Parse both the same way (local midnight) so
          // the comparison lines up exactly with what's shown in the pickers.
          function parseLocalDate(str) {
            var parts = str.split('-');
            return new Date(
              parseInt(parts[0], 10),
              parseInt(parts[1], 10) - 1,
              parseInt(parts[2], 10)
            );
          }

          var startPicker = flatpickr('#explorerStartDate', {
            dateFormat: 'Y-m-d',
            onChange: function(selectedDates) {
              state.startDate = selectedDates.length ? selectedDates[0] : null;
              api.draw();
            }
          });
          var endPicker = flatpickr('#explorerEndDate', {
            dateFormat: 'Y-m-d',
            onChange: function(selectedDates) {
              state.endDate = selectedDates.length ? selectedDates[0] : null;
              api.draw();
            }
          });

          function updateCases() {
            var minVal = $('#explorerCasesMin').val();
            var maxVal = $('#explorerCasesMax').val();
            state.casesMin = minVal === '' ? null : parseFloat(minVal);
            state.casesMax = maxVal === '' ? null : parseFloat(maxVal);
            api.draw();
          }
          $('#explorerCasesMin, #explorerCasesMax').on('change', updateCases);
          $('#explorerCasesMin, #explorerCasesMax').on('keypress', function(e) {
            if (e.which === 13 || e.keyCode === 13) {   // Enter key
              e.preventDefault();
              updateCases();
            }
          });

          // ── Refresh button: clears the calendars, the Cases min/max fields,
          // any table search, and redraws the full, unfiltered dataset.
          $('#explorerRefreshBtn').on('click', function() {
            startPicker.clear();
            endPicker.clear();
            $('#explorerCasesMin').val('');
            $('#explorerCasesMax').val('');
            state.startDate = null;
            state.endDate   = null;
            state.casesMin  = null;
            state.casesMax  = null;
            api.search('').columns().search('').draw();
          });

          $.fn.dataTable.ext.search.push(function(settings2, searchData, index, rowData) {
            if (settings2.nTable.id !== tableId) return true;
            // NOTE: searchData excludes columns marked searchable:false (Date, Cases),
            // so we read from rowData instead — the full, original row values.

            var d = parseLocalDate((rowData[0] || '').toString().trim());   // Date column
            if (state.startDate && d < state.startDate) return false;
            if (state.endDate && d > state.endDate) return false;

            // NOTE: rowData[3] can legitimately be 0, and `0 || ''` evaluates
            // to '' because 0 is falsy in JS — that silently turned every
            // 0-case row's value into NaN, and NaN < min / NaN > max are both
            // false, so 0-case rows always passed the Min/Max filter no
            // matter what was typed. Check for null/undefined explicitly
            // instead of relying on truthiness.
            var rawCases = rowData[3];
            var cs = parseFloat(
              (rawCases === null || rawCases === undefined ? '' : rawCases)
                .toString()
                .replace(/,/g, '')
            );
            if (state.casesMin !== null && cs < state.casesMin) return false;
            if (state.casesMax !== null && cs > state.casesMax) return false;

            return true;
          });
        }
      ")
    ),
    extensions = c("Buttons"),
    class      = "compact stripe hover"
  ) |>
  formatRound("Cases", digits = 0, mark = ",")

htmltools::browsable(htmltools::tagList(date_filter_ui, explorer_table))
```

# Data Sources

## Row

```{=html}
<div style="max-width: 720px; margin: 2rem auto; line-height: 1.7; font-size: 0.95em;">

<p><strong>Dengue case data</strong><br>
Talagala T (2026). <em>denguedatahub: A Tidy Format Datasets of Dengue by Country</em>. R package version 4.1.1. <a href="https://thiyangt.github.io/denguedatahubweb/" target="_blank" rel="noopener">https://thiyangt.github.io/denguedatahubweb/</a></p>

<p><strong>Basemap tiles</strong> (district map, Overview page)<br>
Map tiles by <a href="https://carto.com/attribution/" target="_blank" rel="noopener">CARTO</a>, under CC BY 3.0. Map data by <a href="https://www.openstreetmap.org/copyright" target="_blank" rel="noopener">OpenStreetMap contributors</a>, under the Open Database License (ODbL).</p>

</div>
```
